---
title: "Grouping of RLS sites"
author: "Asta and Freddie"
date: "17 November 2017"
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_float: no
  pdf_document:
    toc:yes
---


```{r warning=FALSE, message=FALSE, warning=FALSE}

rm(list=ls())
list.files()

list.of.packages <- c("tidyverse", "dplyr", "ggplot2", "ggmap")
new.packages     <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)


#load RLS_MPA bassian ecoregion data 
load(file="rlsbas.RData")

nsites <- length(unique(rls_bas$SiteCode))  #531 site codes in the data

## sumarrise bassian sites
sites_bas <- rls_bas %>% 
           group_by(SiteCode) %>% 
           summarise (name = first(`Site name`), location = first(Location), latt = round(mean(SiteLat),3), long = round(mean(SiteLong),3), nsp = n_distinct(SpeciesID), year = n_distinct(year), surveys = n_distinct(SurveyID), depth = mean(Depth), visibility = round(mean(Visibility, na.rm = TRUE),3)) 
# there are 531 sites. 

# Plot the sites
tasmania = get_map(location = c(lon = 145, lat = -40), zoom = 6, maptype = "satellite")

## TODO - add a title and axis names to this 
ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sites_bas$long, y = sites_bas$latt), data = sites_bas, size = 1, color = as.numeric(factor(sites_bas$location))) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))

#Table of sites
#knitr::kable(sites_bas)

```

There are 531 sites. They are roughly grouped into locations as shown on the map, but Rick suggests that location name should not be used. So let's group them differently. First we define a grid of 10 intervals and group the sites into the grid. 

```{r warning=FALSE, message=FALSE, warning=FALSE} 

## Group the sites into 

n_groups <- 10
int <- (max(sites_bas$latt) - min(sites_bas$latt)) / n_groups #latitude interval
sites_bas$group <- ceiling((sites_bas$latt - (min(sites_bas$latt)-0.001)) / int )

int2 <- (max(sites_bas$long) - min(sites_bas$long)) / n_groups #long based interval 
sites_bas$group2 <- ceiling((sites_bas$long - (min(sites_bas$long)-0.001)) / int2 )

sites_bas$group3 <- ((sites_bas$group - 1) * n_groups) + sites_bas$group2  # final grouping, the groups are not consequtive numbers 
length(unique(sites_bas$group3)) # this gives 31 groups


#and map these sites 
hlines <- seq(min(sites_bas$latt),max(sites_bas$latt),by=int)
vlines <- seq(min(sites_bas$long),max(sites_bas$long),by=int2)

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sites_bas$long, y = sites_bas$latt), data = sites_bas, size = 1, color = sites_bas$group3) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0)) +
  geom_vline(xintercept = vlines) +
  geom_hline(yintercept = hlines)

# Now put these groups into the main rls_bas dataset
rls_bas$group <- sites_bas$group3[match(rls_bas$SiteCode, sites_bas$SiteCode)]


```

Next I get site covariate inforamtion. The following chunk is used to load all the data from .csv files and compile it into one data file saved as SiteData_bas.RData. It has a message eval=FALSE which means that it will not run upon knitting. 

```{r warning=FALSE, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE} 

#this chunk only needs to be run once or not at all if the output of site infomation is loaded from RData
#but I keep it here for the future, in case we get new .csv files from Rick and new inforamtion 

#Get other information about the sites, like protection status, exposure, environmental covariates
rls_exp <- read_csv("RLSexposures.csv")
rls_exp <- rls_exp %>% select(SiteCode, `Wave exposure`)
rls_cov <- read_csv("RLSsitecovar.csv") # files with various covariates
rls_cov <- rls_cov[,c(1:2,15:31)]
rls_prot <- read_csv("Neoli.csv")  #file with protection status 
rls_prot <- rls_prot %>% select(SiteCode, Governance)

#in neoli.csv data empty values for governance means that the site is not protected. So we replace them with 0
rls_prot$Governance[is.na(rls_prot$Governance)] <- 0

## These are datafiles from German for the MPA sites
mpa1 <- read_csv("TASprotect1.csv")  #MPA sites and their protection status
mpa2 <- read_csv("TASprotect2.csv")

sites_mpa1 <- mpa1 %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE)) 

a = sapply(sites_mpa1$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_mpa1 <- cbind(sites_mpa1[,1],a)

sites_mpa2 <- mpa2 %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE)) 

b = sapply(sites_mpa2$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_mpa2 <- cbind(sites_mpa2[,1],b)

algae <- read_csv("TASalgae.csv")  #more MPA sites and protection status as well as algal abundance
sites_algae <- algae %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE), abundA = mean(TOTAL_NUMBER)) 

c = sapply(sites_algae$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_algae = cbind(sites_algae[,c(1,3)],c)

##Now I need to merge all the matrices 

sites1 <- merge(x = sites_bas, y = rls_exp, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites2 <- merge(x = sites1, y = rls_cov, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites3 <- merge(x = sites2, y = rls_prot, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites4 <- merge(x = sites3, y = rls_mpa1, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites5 <- merge(x = sites4, y = rls_mpa2, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites6 <- merge(x = sites5, y = rls_algae, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 

# further manipulation of data 

# select the four columns that have the protection status for different sites from different files and merge them to remove NAs
protection <- sites6 %>% select(Governance, a, b, c)
mpa = t(apply(protection, 1, sort, na.last = T)) # now the first column has the protection information 
mpa = as.numeric(mpa[,1]) # pick the first column now, if it has NA then the site does not have protection information. Note, 1 means no take, 0.5 means restricted take, and 0 means open to fishing 

sites6$latt <- abs(sites6$latt) #convert lattitude to positive values for PCA analyses

siteData <- cbind(sites6, mpa)
siteData <- siteData[!duplicated(siteData), ] #remove duplicates

save(siteData, file = "SiteData_bas.RData") # to avoid reading all the matrices again 
```

#Now group sites based on environmental data 

```{r warning=FALSE, message=FALSE, warning=FALSE} 

load (file = "SiteData_bas.RData")

# Select which variables should be included in the final analsyes. I don't include visibility because it is missing in many sites 

siteData1 <- siteData %>% select(SiteCode, latt, long, nsp, depth, damax, chlomean, chlomax, phos, silicate, sstmean, sstrange, sstmax, sstmin, chlomin, parmean, salinity, dissox, cloudmean, nitrate, calcite, ph, abundA, `Wave exposure`)

## The data set above can include any range of selected variables. NOTE, the protection status or variable 'mpa' probably should not be included in clustering. We should cluster sites by different variables. Then the models will be run with and without fishing and emergent data compared to sites with and without protection 

plot(siteData1[,-1])
#It is interesting how latitude correlates with phos and silicate so strongly! Looks a big weird. Latitude correlates with sstmean and sstmax but not sstrange. damax (turbidity) correlates strongly with chlomean and chlomax. So it looks that we could reduce the strongly correlated variables 


#in the new selection I exclude strongly correlated variables and also abundA (algal abudnance) and wave exposure due to missing data
siteData1 <- siteData %>% select(SiteCode, latt, long, nsp, depth, chlomean, phos, sstmean, sstrange, sstmax, sstmin, parmean, salinity, dissox, cloudmean, calcite, ph)

plot(siteData1[,-1])

## Freddie, this is a super clumsy code and I am sure this could be done better - I need to turn it into a numeric matrix wtih site names as row names and variables as columns. This is because clustering analyses do not run on lists but need numeric values 

siteForPCA <- matrix(unlist(siteData1),ncol=length(siteData1[,]),byrow=FALSE) #turn the list into matrix
siteForPCAt <- siteForPCA[,-1] #remove the first column which is the site code
storage.mode(siteForPCAt) <- "numeric"
siteForPCAt[is.nan(siteForPCAt)] <- NA  # some NaN still found in the data and will mess up the analyses. Replace them with NA
rownames(siteForPCAt) <- siteData1[,1] #site code as row names
colnames(siteForPCAt) <- colnames(siteData1[,-1]) #variables as column names

# This is now the main data for PCA analysis of sites based on environmental conditions
envirPCA <- siteForPCAt

##Principal component analysis. 
# At this stage we simply remove observations with missing data. There are some methods that can to interpolate missing data and it might be better to use that in the future 

sitePCA = prcomp(na.omit(envirPCA[,-c(1:2)]), center = TRUE, scale. = TRUE, retx = TRUE) #conduct PCA but don't use lat and long as variables. I use scaled nad rotated axis, but it does not make any difference 
summary(sitePCA)
plot(sitePCA)
biplot(sitePCA, cex = 0.5, expand = 1)
plot(sitePCA$x[,1], sitePCA$x[,2])

sitePCA$rotation[,c(1:2)] #check correlation of variables with the first two axis. Strong correlation show they are most important determinants of grouping

pscores = sitePCA$x[,c(1,2)]

siteData2 = merge(x = siteData1, y = pscores, by.x = "SiteCode", by.y = "row.names", all.x = TRUE) 

# TODO - The following is not a very rigorous or automated way to define groups. This done by simple eyeballing of the PCA plot to identify four main groups on PC1 and PC2. This should be done better 
sitesWithGroups <- siteData2 %>% mutate(g = case_when((siteData2$PC1 < 0 & siteData2$PC2 < 0) ~ 1, siteData2$PC1 > 4 ~  2, (siteData2$PC1 < -2 & siteData2$PC2 >0) ~  3))

sitesWithGroups$g[is.na(sitesWithGroups$g)] <- 4 # assign other sites a code of 4. 

## And plot them now - looks pretty cool! Three clear groups based on environmental variables only 
# We just need to change colours so they are better 
tasmania = get_map(location = c(lon = 145, lat = -40), zoom = 6, maptype = "satellite")

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sitesWithGroups$long, y = -(sitesWithGroups$latt)), data = sitesWithGroups, size = 0.7, color = sitesWithGroups$g) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))


#Finally add the envirogroups to the main dataset
rls_bas$envgroup <- sitesWithGroups$g[match(rls_bas$SiteCode, sitesWithGroups$SiteCode)]


```

The results above show that when 14 variables are included, the first two PC components explain 73% of variation, which is good. The PCA shows that the first axis is mostly determined by the contrast between temperature(sstmean, sstmax), salinity, ph against nitrate, phos, silicate and dissox. The three groups visible on PC1 are further separated on PC2, which is loaded by damax (turbidity),  cholorpyl and sstrage agaist sstmin and depth. The four groups define S, Port Phillip Bay, Derwent and the rest. 

Next we can look whether clustering gives us similar groupings 

```{r warning=FALSE, message=FALSE, warning=FALSE} 
clus <- kmeans(matrixForPCA[,-c(1:2)], 3, nstart = 20) # don't include lat long in clustering, only use environmental covariates

siteclus <- clus$cluster

## add cluster identity to the sizeWithGroups matrix (it is called y now but should be renamed)
sitesWithGroups = merge(x = sitesWithGroups, y = siteclus, by.x = "SiteCode", by.y = "row.names", all.x = TRUE) 

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sitesWithGroups$long, y = -(sitesWithGroups$latt)), data = sitesWithGroups, size = 0.7, color = sitesWithGroups$y) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))
```

*Result* - The k means clustering into 3 groups is not that good. The groups are all intermixed. 

The next thing I want to do is to look at the grouping based on biological data only 


```{r warning=FALSE, message=FALSE, warning=FALSE, eval = FALSE} 

#First filter out species that have only been observed over a certain number of years, here I choose species seen for 5 or more years
# I can either apply a filter by selecting key groups that I think are important, or use all groups

mainSpeciesOccur <- rls_bas  %>% 
 # filter (( PHYLUM == "Chordata" & GENUS != "NA" & CLASS != "Ascidiacea") | CLASS == "Cephalopoda") %>% 
     # filter (( PHYLUM == "Chordata" & GENUS != "NA" & CLASS != "Ascidiacea") | CLASS == "Cephalopoda" | ORDER == "Decapoda" | GENUS == "Haliotis" | GENUS == "Turbo" | CLASS == "Asteroidea" | CLASS == "Echinoidea" | CLASS == "Echiuroidea" ) %>% 
      group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER) %>%
      summarise (years = n_distinct(year), years2000 = n_distinct(ye2000), years2010 = n_distinct(ye2010)) %>%
        filter(years > 4) 

MainSpeciesList <- mainSpeciesOccur[[1]] ## list of 163 species filtered by occurrence in years, or 220 species without any filters

## Common benthic invertebrates - include lobsters (ORDER == "Decapoda", not many of them), abalone (GENUS == "Haliotis", 2 species), periwinkles (GENUS == "Turbo", Turbo undulatus), seastars (Class == "Asteroidea"), and urchins as one grazer group (CLASS == "Echinoidea" & "Echiuroidea"). Also Maoricolpus roseus is a common gastropod and an invader from NZ. Dicathais orbita (sea kangaroo?) is a common gastropod, could be pooled with two other presumably herbivorous gastropods? Alternatively we just pool all gastopods except for Haliotis


#Second calculate mean abundance and biomass of these MORE COMMON species per survey and filter out rare species 
SpPerSurvey <- rls_bas  %>% 
  filter (TAXONOMIC_NAME %in% MainSpeciesList) %>% 
      group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER, SurveyID) %>% ## group by day, month, year and location - survery
        summarise (abundSur = round(sum(Abundance),2), biomSur = round(sum(BioMass, na.rm=TRUE),2)) %>%
          group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER) %>%
            summarise(meanAbun = round(mean(abundSur),3), meanBiom = round(mean(biomSur),3)) %>%
              filter (meanAbun > 4 | meanBiom > 1000)

MainSpecies <- SpPerSurvey[[1]] ## if abund > 5 I get final list of common species with 3 filters applied = 68 species, or 78 species if using all taxa. If abund > 6 I get 68 for final list of taxa. 

#Get summary about the groups 
groups_bas <- rls_bas %>% 
           group_by(group) %>% 
           summarise (location = first(Location), latt = round(mean(SiteLat),3), long = round(mean(SiteLong),3), nsp = n_distinct(SpeciesID), year = n_distinct(year), surveys = n_distinct(SurveyID), locatnum = n_distinct(Location), depth = mean(Depth), visibility = round(mean(Visibility, na.rm = TRUE),3)) 


# Create sitexspecies abundance matrix. First, summarise species records for the main sites. One can apply various filters to exclude locations that have large deviation on sites
SpeciesforPCA <- rls_bas %>%
  filter (TAXONOMIC_NAME %in% MainSpecies) %>%
#    filter (Location != "Port Phillip Bay" & Location != "Port Phillip Heads" & Location != "Rocky Cape" & Location != "Derwent" & Location != "Derwent Estuary" & Location != "Victoria (other)") %>%
 # filter (Location != "Port Phillip Bay" & Location != "Port Phillip Heads") %>%
  filter (SiteLat < -39) %>% # filter victorian sites 
  #    filter(Method == 1) %>%
       #   filter (year < 2006) %>%
            group_by(group, SurveyID, TAXONOMIC_NAME) %>%
                summarise (abund = round(sum(Abundance),3)) %>%
                    group_by(group, TAXONOMIC_NAME) %>%
                        summarise(abundMean = round(mean(abund), 3))

## This gives the mean abundance of key species per site - a large matrix

#Freddie, here again is the same issue of converting a list into a matrix
spPCAdata <- spread(SpeciesforPCA, key = group, value = abundMean, fill = 0)
spPCAdata1 <- matrix(unlist(spPCAdata),ncol=length(spPCAdata[,]),byrow=FALSE) #turn the list into matrix
spPCAdatat <- spPCAdata1[,-1]
storage.mode(spPCAdatat) <- "numeric"
rownames(spPCAdatat) <- spPCAdata1[,1]
colnames(spPCAdatat) <- colnames(spPCAdata[,-1])
speciesPCA <- t(spPCAdatat)

#Freddie, we need to give more informative titles to the groups we now define from the coordinate grid. For example, in the plots below it would be good to identify what those groups are 

## now we do PCA 
sitePCAsp = prcomp(speciesPCA, scale. = TRUE, retx = TRUE) 
summary(sitePCAsp)
plot(sitePCAsp)
biplot(sitePCAsp, cex = 0.5, expand = 1)
plot(sitePCAsp$x[,1], sitePCAsp$x[,2]) # How can we identify the groups better? 

pscores = sitePCAsp$x[,c(1,2)]

groupData = merge(x = groups_bas, y = pscores, by.x = "group", by.y = "row.names", all.x = FALSE) 

# The following is not a very rigorous or automated way to define groups. This done by simple eyeballing of the PCA plot to identify four main groups on PC1 and PC2
groupData <- groupData %>% mutate(grPC = case_when(groupData$PC1 > 2 ~ 1, (groupData$PC1 <1 & groupData$PC2 < -1) ~  2, (groupData$PC1 <1 & groupData$PC2 > -1) ~  3))

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = groupData$long, y = groupData$latt), data = groupData, size = 1, color = (groupData$grPC+1)) + # I add +1 to colour because black colour is almost not visible
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))

```

The last chunk looks at various ordinations based on location 


```{r warning=FALSE, message=FALSE, warning=FALSE, eval = FALSE} 

# The PCA on sites does not look good as there is too much effect of some deviant sites and variation cannot be summarised in the first few axes. Let's try to do the PCA on Locations instaed

#I exclude victorian locations and derwent river as they seem to be too different. Also the analysis can be done for two main study periods - before and after 2005
LocForPCA <- rls_bas %>%
  filter (TAXONOMIC_NAME %in% MainSpecies) %>%
  filter (Location != "Port Phillip Bay" & Location != "Port Phillip Heads" & Location != "Rocky Cape" & Location != "Derwent" & Location != "Derwent Estuary" & Location != "Victoria (other)" & Location != "Tasmania (other)") %>%
    filter (( PHYLUM == "Chordata" & GENUS != "NA" & CLASS != "Ascidiacea") | CLASS == "Cephalopoda") %>% # try grouping on vertebrates only
      filter(Method == 1) %>%
          filter (year > 2005) %>%
            group_by(Location, SurveyID, TAXONOMIC_NAME) %>%
                summarise (abund = round(sum(Abundance),3)) %>%
                    group_by(Location, TAXONOMIC_NAME) %>%
                        summarise(abundMean = round(mean(abund), 3))


locPCAdata <- spread(LocForPCA, key = Location, value = abundMean, fill = 0)
locPCAdata1 <- matrix(unlist(locPCAdata),ncol=length(locPCAdata[,]),byrow=FALSE) #turn the list into matrix
locPCAdatat <- locPCAdata1[,-1]
storage.mode(locPCAdatat) <- "numeric"
rownames(locPCAdatat) <- locPCAdata1[,1]
colnames(locPCAdatat) <- colnames(locPCAdata[,-1])
locationPCA  <-t(locPCAdatat)  ## This is the main biological data for sites


locatPCAsp = prcomp(locationPCA[], scale. = TRUE, retx = TRUE) #conduct PCA on species abundance data only. We need to scale the abundance matrix because it is not correlation matrix and otherwise some common species have a huge effect
summary(locatPCAsp)
plot(locatPCAsp)
biplot(locatPCAsp, cex = c(0.6, 0.5), expand = 1.5, xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.4))  # REMEMBER that the first 2 axes only explain 25% of total variation
plot(locatPCAsp$x[,1], locatPCAsp$x[,3])

locatPCAsp$rotation[,c(1,2)]

## alternatively try MDS on the location abundance data 

#
#mdstest = metaMDS(locationPCA, k=2, trymax = 200)  #  this was first done with most data, but Flinders island affects the data too much

mdstest = metaMDS(locationPCA [-8,], k=2, trymax = 200)  # exclude row 8 which is flinders island 

ordiplot(mdstest,type="n")
orditorp(mdstest,display="sites",cex=1,air=0.1)
orditorp(mdstest,display="species",col="red",air=0.01)

stressplot(mdstest)
plot(mdstest)

## Or alternatively try k-means clustering on locationxspecies data. This is not a very intersting result, because just a few locations are separated and most others are grouped into 1. K-means clustering always separates Ninepin (internal and external) into its own group 

clus <- kmeans(locationPCA, 5, nstart = 20)
locationClusters <- clus$cluster


# Summarise location data
locatSum <- rls_bas %>% 
           group_by(Location) %>% 
           summarise (latt = mean(SiteLat), long = mean(SiteLong), nsp = n_distinct(SpeciesID), year = n_distinct(year), dep_min = min(Depth), dep_max = max(Depth), visib = round(mean(Visibility, na.rm = TRUE),3))


## add cluster identity to the location datda
locationsWithGroups = merge(x = locatSum, y = locationClusters, by.x = "Location", by.y = "row.names", all.x = FALSE) 

## so this clu

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sitesWithGroups$long, y = -(sitesWithGroups$latt)), data = sitesWithGroups, size = 0.7, color = sitesWithGroups$y) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))



## some other code bits and pieces
#temp[, colSums(temp != 0) > 0]
#SelectVar[, colSums(SelectVar == 0) == 0]

```
