---
title: "Grouping of RLS sites"
author: "Asta and Freddie"
date: "17 November 2017"
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_float: no
  pdf_document:
    toc:yes
---



```{r warning=FALSE, message=FALSE, warning=FALSE}

rm(list=ls())
list.files()

list.of.packages <- c("tidyverse", "dplyr", "ggplot2", "ggmap")
new.packages     <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)


#load RLS_MPA bassian ecoregion data 
load(file="rlsbas.RData")

nsites <- length(unique(rls_bas$SiteCode))  #531 site codes in the data

## sumarrise bassian sites
sites_bas <- rls_bas %>% 
           group_by(SiteCode) %>% 
           summarise (name = first(`Site name`), location = first(Location), latt = mean(SiteLat), long = mean(SiteLong), nsp = n_distinct(SpeciesID), year = n_distinct(year), depth = mean(Depth), visibility = round(mean(Visibility, na.rm = TRUE),3)) 
# there are 531 sites. 

# Plot the sites
tasmania = get_map(location = c(lon = 145, lat = -40), zoom = 6, maptype = "satellite")

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sites_bas$long, y = sites_bas$latt), data = sites_bas, size = 1, color = "orange") +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))
```

The following chunk can be ignored for now. It is used to load all the data from .csv files and compile it into one data file saved as SiteData_bas.RData. It has a message eval=FALSE which means that it will not run upon execution

```{r warning=FALSE, message=FALSE, warning=FALSE, eval=FALSE} 

#this chunk only needs to be run once or not at all if the output of site infomation is loaded from RData
#but I keep it here for the future, in case we get new .csv files from Rick and new inforamtion 

#Get other information about the sites, like protection status, exposure, environmental covariates
rls_exp <- read_csv("RLSexposures.csv")
rls_exp <- rls_exp %>% select(SiteCode, `Wave exposure`)
rls_cov <- read_csv("RLSsitecovar.csv") # files with various covariates
rls_cov <- rls_cov[,c(1:2,15:31)]
rls_prot <- read_csv("Neoli.csv")  #file with protection status 
rls_prot <- rls_prot %>% select(SiteCode, Governance)

#in neoli.csv data empty values for governance means that the site is not protected. So we replace them with 0
rls_prot$Governance[is.na(rls_prot$Governance)] <- 0

## These are datafiles from German for the MPA sites
mpa1 <- read_csv("TASprotect1.csv")  #MPA sites and their protection status
mpa2 <- read_csv("TASprotect2.csv")

sites_mpa1 <- mpa1 %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE)) 

a = sapply(sites_mpa1$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_mpa1 <- cbind(sites_mpa1[,1],a)

sites_mpa2 <- mpa2 %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE)) 

b = sapply(sites_mpa2$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_mpa2 <- cbind(sites_mpa2[,1],b)

algae <- read_csv("TASalgae.csv")  #more MPA sites and protection status as well as algal abundance
sites_algae <- algae %>%
  group_by(SiteCode) %>%
    summarise(protection = first(RESERVE_STATUS_CODE), abundA = mean(TOTAL_NUMBER)) 

c = sapply(sites_algae$protection,switch,'NTZ'=1,'EXT'=0, 'RTZ' = 0.5)
rls_algae = cbind(sites_algae[,c(1,3)],c)

##Now I need to merge all the matrices 

sites1 <- merge(x = sites_bas, y = rls_exp, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites2 <- merge(x = sites1, y = rls_cov, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites3 <- merge(x = sites2, y = rls_prot, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites4 <- merge(x = sites3, y = rls_mpa1, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites5 <- merge(x = sites4, y = rls_mpa2, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 
sites6 <- merge(x = sites5, y = rls_algae, by.x = "SiteCode", by.y = "SiteCode", all.x = TRUE) 

# further manipulation of data 

# select the four columns that have the protection status for different sites from different files and merge them to remove NAs
protection <- sites6 %>% select(Governance, a, b, c)
mpa = t(apply(protection, 1, sort, na.last = T)) # now the first column has the protection information 
mpa = as.numeric(mpa[,1]) # pick the first column now, if it has NA then the site does not have protection information. Note, 1 means no take, 0.5 means restricted take, and 0 means open to fishing 

sites6$latt <- abs(sites6$latt) #convert lattitude to positive values for PCA analyses

siteData <- cbind(sites6, mpa)
siteData <- siteData[!duplicated(siteData), ] #remove duplicates

save(siteData, file = "SiteData_bas.RData") # to avoid reading all the matrices again 
```



```{r warning=FALSE, message=FALSE, warning=FALSE} 

load (file = "SiteData_bas.RData")

# Select which variables should be included in the final analsyes. I don't include visibility because it is missing in many sites 

siteData1 <- siteData %>% select(SiteCode, latt, long, nsp, depth, damax, chlomean, chlomax, phos, silicate, sstmean, sstrange, sstmax, sstmin, chlomin, parmean, salinity, dissox, cloudmean, nitrate, calcite, ph, abundA, `Wave exposure`)

## The data set above can include any range of selected variables. NOTE, the protection status or variable 'mpa' probably should not be included in clustering. We should cluster sites by different variables. Then the models will be run with and without fishing and emergent data compared to sites with and without protection 

plot(siteData1)
#It is interesting how latitude correlates with phos and silicate so strongly! Looks a big weird. Latitude correlates with sstmean and sstmax but not sstrange. damax (turbidity) correlates strongly with chlomean and chlomax. So it looks that we could reduce the strongly correlated variables 


#in the new selection I exclude strongly correlated variables and also abundA (algal abudnance) and wave exposure due to missing data
siteData1 <- siteData %>% select(SiteCode, latt, long, nsp, depth, chlomean, phos, sstmean, sstrange, sstmax, sstmin, parmean, salinity, dissox, cloudmean, calcite, ph)

plot(siteData1)


## Freddie, this is a super clumsy code and I am sure this could be done better - I need to turn it into a numeric matrix wtih site names as row names and variables as columns. This is because clustering analyses do not run on lists but need numeric values 

siteForPCA <- matrix(unlist(siteData1),ncol=length(siteData1[,]),byrow=FALSE) #turn the list into matrix
siteForPCAt <- siteForPCA[,-1] #remove the first column which is the site code
storage.mode(siteForPCAt) <- "numeric"
siteForPCAt[is.nan(siteForPCAt)] <- NA  # some NaN still found in the data and will mess up the analyses. Replace them with NA
rownames(siteForPCAt) <- siteData1[,1] #site code as row names
colnames(siteForPCAt) <- colnames(siteData1[,-1]) #variables as column names

# This is now the main data for PCA analysis of sites based on environmental conditions
envirPCA <- siteForPCAt

##Principal component analysis. 
# At this stage we simply remove observations with missing data. There are some methods that can to interpolate missing data and it might be better to use that in the future 

matrixForPCA = na.omit(envirPCA) # seelect variables to be included in analyses - don't include visibility, abundA (algal abundance) and wave exposure as there are too many NAs in those


sitePCA = prcomp(na.omit(matrixForPCA[,-c(1:2)]), center = TRUE, scale. = TRUE, retx = TRUE) #conduct PCA but don't use lat and long as variables. I use scaled nad rotated axis, but it does not make any difference 
summary(sitePCA)
plot(sitePCA)
biplot(sitePCA, cex = 0.5, expand = 1)
plot(sitePCA$x[,1], sitePCA$x[,2])

sitePCA$rotation[,c(1:2)] #check correlation of variables with the first two axis. Strong correlation show they are most important determinants of grouping

pscores = sitePCA$x[,c(1,2)]


```

The results above show that when 14 variables are included, the first two PC components explain 73% of variation, which is good. The PCA shows that the first axis is mostly determined by the contrast between temperature(sstmean, sstmax), salinity, ph against nitrate, phos, silicate and dissox. The three groups visible on PC1 are further separated on PC2, which is loaded by damax (turbidity),  cholorpyl and sstrage agaist sstmin and depth

```{r warning=FALSE, message=FALSE, warning=FALSE} 

#Next I want to assing group membership to sites based on their PC score. At this point I just want to separate sites on the left (PC1 < 0), versus those that have PC1 > 0.05, so I can plot them on a map with different colours. 

siteData2 = merge(x = siteData1, y = pscores, by.x = "SiteCode", by.y = "row.names", all.x = TRUE) 


sitesWithGroups <- siteData2 %>% mutate(g = case_when(siteData2$PC1 < 0 ~ 2, siteData2$PC2 > 0.5 ~  5)) # this could be done better, because I think case_when only works with two conditions?? 

sitesWithGroups <- siteData2 %>% mutate(g = case_when((siteData2$PC1 < 0 & siteData2$PC2 < 0) ~ 1, siteData2$PC1 > 4 ~  2, (siteData2$PC1 < 0 & siteData2$PC2 >0) ~  3)) # this could be done better, because I think case_when only works with two conditions?? 

sitesWithGroups$g[is.na(sitesWithGroups$g)] <- 4 # assign other sites a code of 7. NOTE, these codes are just arbitrary to produce better colours in the plot below, this should be improved. One problem with this approach is that it assigns code to both intermediate group in PCA and also to sites with NA values which were excluded from the analyses

## And plot them now - looks pretty cool! Three clear groups based on environmental variables only 
# We just need to change colours so they are better 
tasmania = get_map(location = c(lon = 145, lat = -40), zoom = 6, maptype = "satellite")

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sitesWithGroups$long, y = -(sitesWithGroups$latt)), data = sitesWithGroups, size = 0.7, color = sitesWithGroups$g) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))

```

Next we can look whether clustering gives us similar groupings 

```{r warning=FALSE, message=FALSE, warning=FALSE} 
clus <- kmeans(matrixForPCA[,-c(1:2)], 3, nstart = 20) # don't include lat long in clustering, only use environmental covariates

siteclus <- clus$cluster

## add cluster identity to the sizeWithGroups matrix (it is called y now but should be renamed)
sitesWithGroups = merge(x = sitesWithGroups, y = siteclus, by.x = "SiteCode", by.y = "row.names", all.x = TRUE) 

ggmap(tasmania, extent = "panel", legend = "bottomright") +
geom_point(aes(x = sitesWithGroups$long, y = -(sitesWithGroups$latt)), data = sitesWithGroups, size = 0.7, color = sitesWithGroups$y) +
scale_y_continuous(limits = c(-44.00, -37.00), expand = c(0, 0))  +
scale_x_continuous(limits = c(143.00, 150.00), expand = c(0, 0))
```

*Result* - The k means clustering into 3 groups gives a little bit similar grouping but not as clearcut. So PCA seems to be better. We need to change colours to make it better visible 

The next thing I want to do is to look at the grouping based on biological data only 


```{r warning=FALSE, message=FALSE, warning=FALSE, eval = FALSE} 

#First filter out species that have only been observed over a certain number of years, here I choose species seen for 5 or more years
# I can either apply a filter by selecting key groups that I think are important, or use all groups

mainSpeciesOccur <- rls_bas  %>% 
  filter (( PHYLUM == "Chordata" & GENUS != "NA" & CLASS != "Ascidiacea") | CLASS == "Cephalopoda") %>% 
     # filter (( PHYLUM == "Chordata" & GENUS != "NA" & CLASS != "Ascidiacea") | CLASS == "Cephalopoda" | ORDER == "Decapoda" | GENUS == "Haliotis" | GENUS == "Turbo" | CLASS == "Asteroidea" | CLASS == "Echinoidea" | CLASS == "Echiuroidea" ) %>% 
      group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER) %>%
      summarise (years = n_distinct(year), years2000 = n_distinct(ye2000), years2010 = n_distinct(ye2010)) %>%
        filter(years > 4) 

MainSpeciesList <- mainSpeciesOccur[[1]] ## list of 163 species filtered by occurrence in years, or 220 species without any filters


## Common benthic invertebrates - include lobsters (ORDER == "Decapoda", not many of them), abalone (GENUS == "Haliotis", 2 species), periwinkles (GENUS == "Turbo", Turbo undulatus), seastars (Class == "Asteroidea"), and urchins as one grazer group (CLASS == "Echinoidea" & "Echiuroidea"). Also Maoricolpus roseus is a common gastropod and an invader from NZ. Dicathais orbita (sea kangaroo?) is a common gastropod, could be pooled with two other presumably herbivorous gastropods? Alternatively we just pool all gastopods except for Haliotis


#Second calculate mean abundance and biomass of these MORE COMMON species per survey and filter out rare species 
SpPerSurvey <- rls_bas  %>% 
  filter (TAXONOMIC_NAME %in% MainSpeciesList) %>% 
      group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER, SurveyID) %>% ## group by day, month, year and location - survery
        summarise (abundSur = round(sum(Abundance),2), biomSur = round(sum(BioMass, na.rm=TRUE),2)) %>%
          group_by (TAXONOMIC_NAME, GENUS, CLASS, FAMILY, ORDER) %>%
            summarise(meanAbun = round(mean(abundSur),3), meanBiom = round(mean(biomSur),3)) %>%
              filter (meanAbun > 6 | meanBiom > 1000)

MainSpecies <- SpPerSurvey[[1]] ## if abund > 5 I get final list of common species with 3 filters applied = 68 species, or 78 species if using all taxa. If abund > 6 I get 68 for final list of taxa. 

SpeciesforPCA <- rls_bas %>%
  filter (TAXONOMIC_NAME %in% MainSpecies) %>%
    filter(Method == 1) %>%
#          filter (year > 1999, year <2010) %>%
            group_by(SiteCode, SurveyID, TAXONOMIC_NAME) %>%
                summarise (abund = round(sum(Abundance),3)) %>%
                    group_by(SiteCode, TAXONOMIC_NAME) %>%
                        summarise(abundMean = round(mean(abund), 3))

## This gives the mean abundance of key species per site - a large matrix

spPCAdata <- spread(SpeciesforPCA, key = SiteCode, value = abundMean, fill = 0)
spPCAdata1 <- matrix(unlist(spPCAdata),ncol=length(spPCAdata[,]),byrow=FALSE) #turn the list into matrix
spPCAdatat <- spPCAdata1[,-1]
storage.mode(spPCAdatat) <- "numeric"
rownames(spPCAdatat) <- spPCAdata1[,1]
colnames(spPCAdatat) <- colnames(spPCAdata[,-1])
speciesPCA  <-t(spPCAdatat)  ## This is the main biological data for sites

## now we have speciesPCA and envirPCA data 

#Merge all data 

PCAdataMain <- merge(x = envirPCA, y = speciesPCA, by = "row.names", all.x = TRUE) 
rownames(PCAdataMain) = PCAdataMain[,1] #move first column into the row names
PCAdataMain = PCAdataMain[,-1]

```

The PCAdataMain is the main data file with environmental and species abundance data for 531 sites


```{r warning=FALSE, message=FALSE, warning=FALSE, eval = FALSE} 


sitePCAsp = prcomp(na.omit(PCAdataMain[,-c(1:16)]), scale. = TRUE, retx = TRUE) #conduct PCA on species abundance data only. We need to scale the abundance matrix because it is not correlation matrix and otherwise some common species have a huge effect
summary(sitePCAsp)
plot(sitePCAsp)
biplot(sitePCAsp, cex = 0.5, expand = 1)
plot(sitePCAsp$x[,1], sitePCAsp$x[,2])

# we can see that there are some outlier sites
loadingsp <- sitePCAsp$rotation[,c(1,2)]

spcores = sitePCAsp$x[,c(1,2)]
siteData2 = merge(x = PCAdataMain, y = spcores, by = "row.names", all.x = TRUE) 
rownames(siteData2) = siteData2[,1]
siteData2 = siteData2[,-1]

siteswithSpPC <- siteData2 %>% mutate(g = case_when(siteData2$PC1 < 0 ~ 2, siteData2$PC2 > 0.5 ~  5)) # this could be done better, because I think case_when only works with two conditions?? 
sitesWithGroups$g[is.na(sitesWithGroups$g)] <- 7 # assign other sites a code of 7. NOTE, these codes are just arbitrary to produce better colours in the plot below, this should be improved. One problem with this approach is that it assigns code to both intermediate group in PCA and also to sites with NA values which were excluded from the analyses





temp <- siteData2 %>% filter(siteData2$PC1 < 5 & siteData2$PC2 > -5)  #this removes about 4 deviant sites

temp[, colSums(temp != 0) > 0]


SelectVar[, colSums(SelectVar == 0) == 0]

sitePCAsp2 = prcomp(na.omit(temp[,c(17:69)]), scale. = TRUE, retx = TRUE) 

PCscores= sitePCAsp$x[,c(1:30)] # first 30 PCs include 70% of variation 


mdstest = metaMDS(na.omit(PCAdataMain[,-c(1:16)]), k=2) #no convergence 
#mdstest = metaMDS(dataMDSmt, distance="bray",k=3,trymax=200)
#bas_dist = vegdist(dataMDS)

stressplot(mdstest)
plot(mdstest)

par(mfrow=c(1,1))

ordiplot(mdstest,type="n")
orditorp(mdstest,display="sites",cex=1,air=0.1)
orditorp(mdstest,display="species",col="red",air=0.01)



which(sitePCAsp$x== min(sitePCAsp$x[,1]))

sitePCA$rotation[,c(1:2)] #check correlation of variables with the first two axis. Strong correlation show they are most important determinants of grouping

pscores = sitePCA$x[,c(1,2)]


biolData = na.omit(PCAdataMain[,-c(3:25)])


